Calculating the Solvent Exposed Surface Area of Proteins with Intersecting Spheres

### 2020-06-06 Paper.tex
more changes

Our protein can be represented as a set of atom centers $S$ in $\R^3$ with their corresponding balls of various radii. One aspect of the model is computing the power diagram of this protein. The power diagram is a generalized Voronoi diagram, where instead of the normal Euclidean metric, the definition for distance between an atom with center $z_i$ and radius $\varrho_i$ and a point $x$ is $\pi_i(x)=\|x-z_i\|^2-\varrho_i^2$. This becomes the default Voronoi diagram of $S$ if all the radii are equal. After this power diagram is generated we can associate each atom $S_i$ with its corresponding power cell $P_i$ defined by
\begin{equation*}
P_i=\{x\in\R^3\mid\pi_i(x)\leq\pi_j(x),\forall j\}
\end{equation*}
Each $P_i$ is convex, and together they cover all $\R^3$ without overlap. The dual of this power diagram is a weighted Delaunay Triangulation of $S$ into a set of simplices. We can compute the surface area of these balls using Delaunay triangulation, which will help us in the later sections. There are additional rules for constructing Delaunay triangles, but for the purposes of this paper, it is only necessary to how they are constructed from the atoms.

@@

Consider a set of balls $B_i$ defined by a center $z_i$ and a fixed radius $\varrho_i$ to be our protein. To represent the entire state of the protein (with $n$ atoms) we use the $m=3n$ dimensional  vector $\textbf{z}\in\R^{3n}$ where $\lbrack{\bf z}_{3i+1}, {\bf z}_{3i+2}, {\bf z}_{3i+3}\rbrack^T=z_i$. Now we can define the weighted and unweighted surface areas $E,A:\R^{3m}\rightarrow\R$ as real multivariate functions from the state of the protein to a single value (the solvent exposed surface area). The weighted area will be described below. Here we will describe the gradients of these maps, written as $DE_{\bf z}$ and $DA_{\bf z}$ which would allow us to calculate SASA's for protein structures and approximate how the SASA changes as the atomic coordinates move. These derivatives map each point in $\R^m$ to a corresponding gradient vector that we call \textbf{u}.

@@

In order to derive the SASA equations, we must define some additional terms. After constructing Delaunay Triangles from a set of molecules, the “star” and “link” of the alpha shape's subcomplex are defined. Each of the Delaunay triangles as a simplex, where simplex is basically refers to a triangle or tetrahedron. K is defined as a dual complex, or subcomplex of the alpha shape.

\begin{figure}[h!]
\caption{The dual complex (K) of the Delaunay Triangulation only includes those power cells that have a non-empty intersection.$\lbrack1\rbrack$}
\centerline{\includegraphics[width=3in]{Figure9}}
\end{figure}

As can be seen by Figure 6, K is made up of multiple simplices. The star of K is a set of triangles who share a vertex and outside boundary. The link of K, then, is all of the vertices and edges in the star (Figure 7).

\begin{figure}[h!]
\caption{The star of K and link of the star are shown.$\lbrack1\rbrack$}
\centerline{\includegraphics[width=3in]{Figure5}}
\end{figure}

Knowing how the star and link of a subcomplex K are defined is useful when deriving the SASA equations.
The collection of spheres, or the entire molecule, is known as a space-filling diagram $F$. $F_i$, for instance, consists of all the atoms in $S_i$. $\sigma$ represents the fraction of a sphere on the boundary of $F$. We only really care about the boundary of $F$ because that is what contributes to the exposed surface area that will be interacting with the solvent. As mentioned earlier, a molecule can contain multiple spheres that overlap. For two intesecting spheres $S_i$ and $S_j$ that share a circle $S_{ij}$ $\sigma$ comes out to:

\begin{equation*}
\sigma_{ij} = \frac{length(P_i \cap P_j \cap bd F)}{length(S_{ij})}
\end{equation*}

@@

\begin{equation*}
\sigma_{ij} = 1 - \frac{\sum_{k}length(S^k_{ij})- \sum_{k,l}length(S^k_{ij} \cap S^l_{ij})}{2\pi\varrho_{ij}}
\end{equation*}

This equation is derived by looking at Edelsbrunner's (1995) paper. The same two spheres $S_i$ and $S_j$ intersect at $S_{ij}$. However this time, $S^k_{ij}$ and $S^l_{ij}$ are two other spheres in the power diagram coming into contact with $S_{ij}$. Both  $S^k_{ij}$ and $S^l_{ij}$ minimize the power distance to $S_{ij}$. Looking at the term on the right of equation 7, the first sum adds the power distances from the portion of $S_{ij}$ coming into contact with $S_k$, while the second sum adds the power distances from the portion of $S_{ij}$ coming into contact with both $S_k$ and $S_l$. Having this second sum ensures we don't underestimate the amount of unexposed area. Dividing the difference of these sums by the circumference of the circle gives the fraction of this unexposed area, so that subtracting this term from 1 gives the fraction of exposed surface area of $F$.

$\sigma$ is related to the area of the ball, which for a ball with radius $\varrho$ is $4\pi\varrho^2$. Because we only care about the exposed surface area, we need to multiply by $\sigma$, in order to only get the solvent-exposed surface area. We can also define a weighted area, which only differs from the area equation by the factor of some parameter $\alpha$, which represents some constant atomic solvation parameters that were predefined by Eisenberg and McLachlan (1986). So for just one ball $B_i$:
\begin{align*}
&A = 4\pi\sum \varrho_i^2\sigma_i
&E = 4\pi\sum\alpha_i\varrho_i^2\sigma_i
\end{align*}

Another term, $\beta_{ijk}$, is also used, which is useful when considering the case of three intersecting balls. When three spheres intersect at two points, there is a line segment passing through them that is represented as $B_{ijk}$. Just like for $\sigma_{ij}$, $\beta_{ijk}$ represents the fraction of points that belong to the edge of $F$ for these three spheres.
\begin{equation*}
\beta_{ijk} = \frac{length(P_i \cap P_j \cap P_k \cap bd F)}{length(B_{ijk})}
\end{equation*}

A similar expression to $\sigma_{ij}$ is derived for $\beta_{ijk}$, using $B^l_{ijk}$, the portion of the line segment $B_{ijk}$ on a fourth sphere $S_l$. 
\begin{equation*}
\beta_{ijk} = 1 - \frac{\sum_{l}length(B^l_{ijk})}{length(B_{ijk})}
\end{equation*}
In the right term, the sum adds up the length of the line segment that comes in contact with $S_l$ and divides that by the entire length of $B_{ijk}$ (which is $2\varrho_{ijk}$), giving the fraction of $F$ that is unexposed. By subtracting this term from 1, the fraction of solvent exposed area for the three intersecting spheres is found.

We can think of the sphere as representing how much of the solvent can interact with the atom. 
\begin{comment}
The state \textbf{z} of each ball defines a center and a radius. As stated before, this paper considers that the weighted (E) and unweighted (A) areas are transformations from the space containing z ($\R^{3n}$) to $\R$. Their derivatives at a particular \textbf{z} are linear maps. In other words, the areas are represented by the map f and their derivatives the derivative of f at \textbf{z}.
\end{comment}
Because the area derivatives are linear, they can be broken down into a sum of components. Both unweighted and weighted area derivatives are comprised of two motions, direction preserving and distance preserving, which can be derived separately and summed to form the theorems.

@@

The general equation of a sphere ($x^2 + y^2 +z^2 = r^2$) can be used to construct equations for $S_i$ and $S_j$:
\begin{align*}
&x^2 + y^2 +(z-z_i)^2 = \zeta_i^2 \\
&x^2 + y^2 +(z-z_j)^2 = \zeta_j^2 \\
&\zeta_i^2-\zeta_j^2 =(z-z_i)^2-(z-z_j)^2=-2z_i +2z_j +z_i^2 - z_j^2\\
& \implies -2z_i +2z_j +z_i^2 - z_j^2 = (\zeta_i)^2-(\zeta_j)^2\\
\end{align*}
Using this logic, we can find the same equation holds true if we consider the set of balls for each sphere (assuming they have they same radii), as they have the same center as the spheres. So:
\begin{align*}
&-2z_i +2z_j +z_i^2 - z_j^2 = \varrho_i^2-\varrho_j^2 \\
&\text{From the above two expressions, we can state that:} \\
&\zeta_i^2-\zeta_j^2 = \varrho_i^2-\varrho_j^2\\
&\text{In the left term, we can factor out $\zeta_i^2-\zeta_j^2$ to get:} \\
&\zeta_i^2-\zeta_j^2 =(\zeta_i-\zeta_j)(\zeta_i+\zeta_j) = \zeta_{ij}(\zeta_i-\zeta_j) \\
&\text{So we can state that:} \\
&\varrho_i^2 - \varrho_j^2  = \zeta_{ij}(\zeta_i-\zeta_j) \\
&\text{Now we can find what $\zeta_i$ and $\zeta_j$ really are} \\
&\varrho_i^2 - \varrho_j^2)/\zeta_{ij} + \zeta_j = \zeta_i \\
&\text{We know that $\zeta_j = \zeta_{ij} - \zeta_i$, so just substitute this expression in:} \\
&\zeta_i = \frac{1}{2}\left(\zeta_{ij}+\frac{\varrho_i^2-\varrho_j^2}{\zeta_{ij}}\right) \\
&\text{A similar statement can be made for $\zeta_j$, following the steps we carried out above:} \\
&\zeta_j = \frac{1}{2}\left[\zeta_{ij}+\frac{\varrho_j^2-\varrho_i^2}{\zeta_{ij}}\right]
\end{align*}

Now we can find the area of both spheres. For two spheres, it makes sense that $A = A_i + A_j$. Recall the area equation stated in Section 3.3. For two balls we can say: $A = 4\pi\varrho^2\sum{\sigma_i} + 4\pi\varrho^2\sum{\sigma_j}$. We have already shown what $\sigma$ is, so we can plug it in and simplify the area equation:
\begin{align*}
&4\pi\varrho^2\frac{\varrho_i + \zeta_i}{2\varrho_i} + 4\pi\varrho_j^2\frac{\varrho_j + \zeta_j}{2\varrho_j} \\
&\text{After a few algebraic simplifications:} \\
&2\pi\varrho_i^2 +2\pi\varrho_j^2 +\pi\varrho_i\left(\zeta_{ij} + \frac{\varrho_i^2 - \varrho_j^2}{\zeta_{ij}}\right)+\pi\varrho_j\left(\zeta_{ij} + \frac{\varrho_j^2 - \varrho_i^2}{\zeta_{ij}}\right) \\
&A  = 2\pi(\varrho_i^2 + \varrho_j^2) + \pi(\varrho_i+\varrho_j)\left[\zeta_{ij}+\frac{(\varrho_i-\varrho_j)^2}{\zeta_{ij}}\right] \\
&\text{Finally, we can find the derivative of the area with respect to $\zeta_{ij}$:} \\
&\frac{dA}{d\zeta_{ij}} = \pi(\varrho_i+\varrho_j)\left[1-\frac{(\varrho_i - \varrho_j)^2}{\zeta_{ij}^2}\right] \\
&\text{This is term we get after some simplification.}
\end{align*}

@@

Using this spherical rectangle shape, and the fact that we know the caps are rotating, the change in area is found to be the change in the difference between the center of the circles $\eta_j$ divided by the circumference of the set of balls in the larger sphere (which is $S_j$ in this case) times the area of the slice of $S_j$ between $p$ and $q$: 
\begin{align}
dA_j = \frac{d\eta_j}{2\pi\varrho_j}4\pi\varrho_j\varrho_{ijk}\\
\end{align}
$\frac{d\eta_j}{2\pi\varrho_j}$ acts like the change in the “width” of the spherical rectangle while $4\pi\varrho_j\varrho_{ijk}$ serves as the change in the “height” of the spherical rectangle. The area derivative of this larger ball becomes:
\begin{equation*}
\frac{dA_j}{d\eta_j }= 2\varrho_{ijk}
\end{equation*}

@@

The first sum shows the contribution of the direction preserving motion while the second sum shows the contribution of the distance preserving motion. We can see for the first sum that we are summing over all of the points along $F$ when considering the intersection of two spheres $S_i$ and $S_j$, and the second sum considers the sum of points along the intersection for three intersecting spheres. From this, $a_{ij}$ and $a_{ijk}$ are:
\begin{align*}
&a_{ij} = \pi(\varrho_i+\varrho_j)\left[1 - \frac{(\varrho_i - \varrho_j)^2}{\zeta_{ij}^2}\right]\cdot u_{ij}\\
&a_{ijk} = 2\varrho_{ijk}\frac{\varrho_i - \varrho_j}{\zeta_{ij}} \cdot u_{ijk}
\end{align*}
$a_{ij}$ and $a_{ijk}$ depend on the unit vectors $u_{ij}$ and $u_{ijk}$, respectively.

The weighted area derivative theorem, defined as $DE_z(\textbf{t}) = <\textbf{e},\textbf{t}>$ is defined similarly to the area derivative theorem, but multiplied by the solvent parameter $\alpha$:

@@

Above, we described $\textbf{a}$ and $\textbf{e}$. Using these, we can then clearly state the area derivative equations. For the unweighted area derivative equation, the direction preserving component is the the fraction of the solvent-exposed surface times the derivative of the area: $\sigma_{ij}<a_{ij},t_i>$ and the distance preserving component is the same: $\beta_{ijk}<a_{ijk},t_i>$. We know that the unweighted area derivative is the sum of these two components:
\begin{equation}
DA_z(\textbf{t}) = \sum_{i=0}^{n-1}\sum_{j\neq i}\left(\sigma_{ij}<a_{ij},t_i> + \sum_{k\neq i,j} \beta_{ijk}<a_{ijk},t_i>\right)
\end{equation}
Just like before, the weighted area derivative is basically the same as the unweighted derivative: 
\begin{equation}
DE_z(\textbf{t}) = \sum_{i=0}^{n-1}\sum_{j\neq i}\left(\sigma_{ij}<e_{ij},t_i> + \sum_{k\neq i,j} \beta_{ijk}<e_{ijk},t_i>\right)
\end{equation}
These are the derivative equations that can then be implemented into a software and run on different molecules. It is important to note that there are cases where these areas will be discontinuous. We won't elaborate on these, but, for example, discontinuities can happen in a case where three spheres intersect at a circle. There are other cases of discontinuities, which are useful to know in order successfully to run this algorithm for a variety of molecules.

>>>>

Our protein can be represented as a set of atom centers $S$ in $\R^3$ with their corresponding balls of various radii. One aspect of the model is computing the power diagram of this protein. The power diagram is a generalized Voronoi diagram, where instead of the normal Euclidean metric, the definition for distance between an atom with center $z_i$ and radius $\varrho_i$ and a point $x$ is $\pi_i(x)=\|x-z_i\|^2-\varrho_i^2$. This becomes the default Voronoi diagram of $S$ if all the radii are equal. After this power diagram is generated we can associate each set of atoms $S_i$ with its corresponding power cell $P_i$ defined by
\begin{equation*}
P_i=\{x\in\R^3\mid\pi_i(x)\leq\pi_j(x),\forall j\}
\end{equation*}
Each $P_i$ is convex, and together they cover all $\R^3$ without overlap. The dual of this power diagram is a weighted Delaunay Triangulation of $S$ into a set of simplices, or triangles and tetrahedrons. We can compute the surface area of these balls using Delaunay triangulation, which will help us in the later sections. There are additional rules for constructing Delaunay triangles, but for the purposes of this paper, it is only necessary to know how they are constructed from the atoms.

@@

Consider a set of balls $B_i$ defined by a center $z_i$ and a fixed radius $\varrho_i$ to be our protein. To represent the entire state of the protein (with $n$ atoms) we use the $m=3n$ dimensional  vector $\textbf{z}\in\R^{3n}$ where $\lbrack{\bf z}_{3i+1}, {\bf z}_{3i+2}, {\bf z}_{3i+3}\rbrack^T=z_i$. Now we can define the weighted and unweighted surface areas $E,A:\R^{3m}\rightarrow\R$ as real multivariate functions from the state of the protein to a single value (the solvent exposed surface area). The weighted area will be described below. Here we will describe the gradients of these maps, written as $DE_{\bf z}$ and $DA_{\bf z}$ which would allow us to calculate SASAs for protein structures and approximate how the SASA changes as the atomic coordinates move. These derivatives map each point in $\R^m$ to a corresponding gradient vector that we call \textbf{u}.

@@

In order to derive the SASA equations, we must define some additional terms. After constructing Delaunay Triangles from a set of molecules, the “star” and “link” of the alpha shape's subcomplex are defined. Each of the Delaunay triangles is a simplex. $K$ is defined as a dual complex, or subcomplex of the alpha shape.

\begin{figure}[h!]
\caption{The dual complex ($K$) of the Delaunay Triangulation only includes those power cells that have a non-empty intersection.$\lbrack1\rbrack$}
\centerline{\includegraphics[width=3in]{Figure9}}
\end{figure}

As can be seen by Figure 6, $K$ is made up of multiple simplices. The star of $K$ is a set of triangles who share a vertex and outside boundary. The link of $K$, then, is all of the vertices and edges in the star (Figure 7).

\begin{figure}[h!]
\caption{The star of $K$ and link of the star are shown.$\lbrack1\rbrack$}
\centerline{\includegraphics[width=3in]{Figure5}}
\end{figure}

Knowing how the star and link of a subcomplex $K$ are defined is useful when deriving the SASA equations.
The collection of spheres, or the entire molecule, is known as a space-filling diagram $F$. $F_i$, for instance, consists of all the atoms in $S_i$. $\sigma$ represents the fraction of a sphere on the boundary of $F$. We only really care about the boundary of $F$ because that is what contributes to the exposed surface area that will be interacting with the solvent. As mentioned earlier, a molecule can contain multiple spheres that overlap. For two intesecting spheres $S_i$ and $S_j$ that share a circle $S_{ij}$, $\sigma$ comes out to:

\begin{equation}
\sigma_{ij} = \frac{length(P_i \cap P_j \cap bd F)}{length(S_{ij})}
\end{equation}

@@

\begin{equation}
\sigma_{ij} = 1 - \frac{\sum_{k}length(S^k_{ij})- \sum_{k,l}length(S^k_{ij} \cap S^l_{ij})}{2\pi\varrho_{ij}}
\end{equation}

This equation is derived by looking at Edelsbrunner's (1995) paper. The same two spheres $S_i$ and $S_j$ intersect at $S_{ij}$. However this time, $S^k_{ij}$ and $S^l_{ij}$ are the interesction of $S_{ij}$ with two other spheres in the power diagram. Both  $S^k_{ij}$ and $S^l_{ij}$ minimize the power distance to $S_{ij}$. Looking at the right side term of equation 7, the first sum adds the portion of $S_{ij}$ coming into contact with $S_k$, while the second sum adds the portion of $S_{ij}$ coming into contact with both $S_k$ and $S_l$. Having this second sum ensures we don't underestimate the amount of unexposed area. Dividing the difference of these sums by the circumference of the circle gives the fraction of this unexposed area, so that subtracting this term from 1 gives the fraction of the exposed surface area of $F$.

$\sigma$ is related to the area of the ball, which for a ball with radius $\varrho$ is $4\pi\varrho^2$. Because we only care about the exposed surface area, we need to multiply by $\sigma$, in order to only get the solvent-exposed surface area. We can also define a weighted area, which only differs from the unweighted area equation by the factor of some parameter $\alpha$, which represents some constant atomic solvation parameters that were predefined by Eisenberg and McLachlan (1986). So for just one ball $B_i$:
\begin{align*}
&A = 4\pi\sum \varrho_i^2\sigma_i
&E = 4\pi\sum\alpha_i\varrho_i^2\sigma_i
\end{align*}

Another term, $\beta_{ijk}$, is also used, which is useful when considering the case of three intersecting spheres. When three spheres intersect at two points, there is a line segment passing through them that is represented as $B_{ijk}$. Just like for $\sigma_{ij}$, $\beta_{ijk}$ represents the fraction of points that belong to the edge of $F$ for these three spheres.
\begin{equation}
\beta_{ijk} = \frac{length(P_i \cap P_j \cap P_k \cap bd F)}{length(B_{ijk})}
\end{equation}

A similar expression to $\sigma_{ij}$ is derived for $\beta_{ijk}$, using $B^l_{ijk}$, the portion of the line segment $B_{ijk}$ on a fourth sphere $S_l$. 
\begin{equation}
\beta_{ijk} = 1 - \frac{\sum_{l}length(B^l_{ijk})}{length(B_{ijk})}
\end{equation}
In the right side term of equation 9, the sum adds up the length of the line segment that comes in contact with $S_l$ and divides that by the entire length of $B_{ijk}$ (which is $2\varrho_{ijk}$), giving the fraction of $F$ that is unexposed. By subtracting this term from 1, the fraction of solvent exposed area for the three intersecting spheres is found.

We can think of the sphere as representing how much of the solvent can interact with the atom. 
\begin{comment}
The state \textbf{z} of each ball defines a center and a radius. As stated before, this paper considers that the weighted (E) and unweighted (A) areas are transformations from the space containing z ($\R^{3n}$) to $\R$. Their derivatives at a particular \textbf{z} are linear maps. In other words, the areas are represented by the map f and their derivatives the derivative of f at \textbf{z}.
\end{comment}
Because the area derivatives are linear, they can be broken down into a sum of components. Both unweighted and weighted area derivatives are comprised of two motions, direction preserving and distance preserving, which can be derived separately and summed to get the SASA derivatives.

@@

The general equation of a sphere ($x^2 + y^2 +z^2 = r^2$) can be used to construct equations for $S_i$ and $S_j$:
\begin{align*}
&x^2 + y^2 +(z-z_i)^2 = \zeta_i^2 \\
&x^2 + y^2 +(z-z_j)^2 = \zeta_j^2 \\
&\text{Now we can combine the equations:} \\
&\zeta_i^2-\zeta_j^2 =(z-z_i)^2-(z-z_j)^2=-2z_i +2z_j +z_i^2 - z_j^2\\
& \implies -2z_i +2z_j +z_i^2 - z_j^2 = (\zeta_i)^2-(\zeta_j)^2\\
\end{align*}
Using this logic, we can find the same equation holds true if we consider the set of balls for each sphere (assuming they have they same radii), as they have the same center as the spheres. So:
\begin{align*}
&-2z_i +2z_j +z_i^2 - z_j^2 = \varrho_i^2-\varrho_j^2 \\
&\text{From the above two expressions, we can state that:} \\
&\zeta_i^2-\zeta_j^2 = \varrho_i^2-\varrho_j^2\\
&\text{In the left term, we can factor out $\zeta_i^2-\zeta_j^2$ to get:} \\
&\zeta_i^2-\zeta_j^2 =(\zeta_i-\zeta_j)(\zeta_i+\zeta_j) = \zeta_{ij}(\zeta_i-\zeta_j) \\
&\text{So we can state that:} \\
&\varrho_i^2 - \varrho_j^2  = \zeta_{ij}(\zeta_i-\zeta_j) \\
&\text{Now we can find what $\zeta_i$ and $\zeta_j$ really are} \\
&\varrho_i^2 - \varrho_j^2)/\zeta_{ij} + \zeta_j = \zeta_i \\
&\text{We know that $\zeta_j = \zeta_{ij} - \zeta_i$, so just substitute this expression in:} \\
&\zeta_i = \frac{1}{2}\left(\zeta_{ij}+\frac{\varrho_i^2-\varrho_j^2}{\zeta_{ij}}\right) \\
&\text{A similar statement can be made for $\zeta_j$, following the steps we carried out above:} \\
&\zeta_j = \frac{1}{2}\left[\zeta_{ij}+\frac{\varrho_j^2-\varrho_i^2}{\zeta_{ij}}\right]
\end{align*}

Now we can find the area of both spheres. For two spheres, it makes sense that $A = A_i + A_j$. Recall the area equation stated in Section 3.3. For two balls we can say: $A = 4\pi\varrho^2\sum{\sigma_i} + 4\pi\varrho^2\sum{\sigma_j}$. We have already shown what $\sigma$ is, so we can plug it in and simplify the area equation:
\begin{align*}
&4\pi\varrho_i^2\frac{\varrho_i + \zeta_i}{2\varrho_i} + 4\pi\varrho_j^2\frac{\varrho_j + \zeta_j}{2\varrho_j} \\
&\text{After a few algebraic simplifications:} \\
&2\pi\varrho_i^2 +2\pi\varrho_j^2 +\pi\varrho_i\left(\zeta_{ij} + \frac{\varrho_i^2 - \varrho_j^2}{\zeta_{ij}}\right)+\pi\varrho_j\left(\zeta_{ij} + \frac{\varrho_j^2 - \varrho_i^2}{\zeta_{ij}}\right) \\
&A  = 2\pi(\varrho_i^2 + \varrho_j^2) + \pi(\varrho_i+\varrho_j)\left[\zeta_{ij}+\frac{(\varrho_i-\varrho_j)^2}{\zeta_{ij}}\right] \\
&\text{Finally, we can find the derivative of the area with respect to $\zeta_{ij}$:} \\
&\frac{dA}{d\zeta_{ij}} = \pi(\varrho_i+\varrho_j)\left[1-\frac{(\varrho_i - \varrho_j)^2}{\zeta_{ij}^2}\right] \\
&\text{This is term we get after some simplification.}
\end{align*}

@@

Using this spherical rectangle shape, and the fact that we know the caps are rotating, the change in area is found to be the change in the distance between the center of the circles $\eta_j$ divided by the circumference of the set of balls in the larger sphere (which is $S_j$ in this case) times the area of the slice of $S_j$ between $p$ and $q$: 
\begin{align*}
dA_j = \frac{d\eta_j}{2\pi\varrho_j}4\pi\varrho_j\varrho_{ijk}\\
\end{align*}
$\frac{d\eta_j}{2\pi\varrho_j}$ acts like the change in the “width” of the spherical rectangle while $4\pi\varrho_j\varrho_{ijk}$ serves as the change in the “height” of the spherical rectangle. The area derivative of this larger sphere becomes:
\begin{equation*}
\frac{dA_j}{d\eta_j }= 2\varrho_{ijk}
\end{equation*}

@@

The first sum of equation 10 shows the contribution of the direction preserving motion while the second sum shows the contribution of the distance preserving motion. We can see for the first sum that we are summing over all of the points along $F$ in $K$ when considering direction preserving motion for $S_i$ across all $S_j$. Similarly, the second sum considers distance preserving motion for $S_i$ across all $S_j$ and $S_k$ in $K$. From this, $a_{ij}$ and $a_{ijk}$ are:
\begin{align*}
&a_{ij} = \pi(\varrho_i+\varrho_j)\left[1 - \frac{(\varrho_i - \varrho_j)^2}{\zeta_{ij}^2}\right]\cdot u_{ij}\\
&a_{ijk} = 2\varrho_{ijk}\frac{\varrho_i - \varrho_j}{\zeta_{ij}} \cdot u_{ijk}
\end{align*}
$a_{ij}$ and $a_{ijk}$ depend on the unit vectors $u_{ij}$ and $u_{ijk}$, respectively.

The weighted area derivative theorem, defined as $DE_z(\textbf{t}) = <\textbf{e},\textbf{t}>$ is defined similarly to the unweighted area derivative theorem, but multiplied by the solvent parameter $\alpha$:

@@

Above, we described $\textbf{a}$ and $\textbf{e}$. Using these, we can then clearly state the area derivative equations. For the unweighted area derivative equation, the direction preserving component is the the fraction of the solvent-exposed surface times the derivative of the area: $\sigma_{ij}<a_{ij},t_i>$ and the distance preserving component is the same: $\beta_{ijk}<a_{ijk},t_i>$. We know that the unweighted area derivative is the sum of these two components:
\begin{equation}
DA_z(\textbf{t}) = \sum_{i=0}^{n-1}\sum_{j\neq i}\left(\sigma_{ij}<a_{ij},t_i> + \sum_{k\neq i,j} \beta_{ijk}<a_{ijk},t_i>\right)
\end{equation}
Just like before, the weighted area derivative is basically the same as the unweighted derivative: 
\begin{equation}
DE_z(\textbf{t}) = \sum_{i=0}^{n-1}\sum_{j\neq i}\left(\sigma_{ij}<e_{ij},t_i> + \sum_{k\neq i,j} \beta_{ijk}<e_{ijk},t_i>\right)
\end{equation}
These are the derivative equations that can then be tested on different molecules. It is important to note that there are cases where these equations will be discontinuous. We won't elaborate on these, but, for example, discontinuities can happen in a case where three spheres intersect at a circle. There are other cases of discontinuities, which are useful to know in order successfully to run this algorithm for a variety of molecules.
